The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2020-11-04 Manuscript/NBABubbleAnalysis.Rmd
Update NBABubbleAnalysis.Rmd

---
title: "The Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price and Jun Yan
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: bookdown::pdf_document2
header-includes:
   - \usepackage{float}
   - \usepackage[margin=1in]{geometry}
toc: False
---
Abstract: 
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there would be no fans in attendance, 
games played on neutral courts and no traveling for teams, which in theory 
removes home-court advantage from the games. This setting has attracted a lot of
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played a huge role in 
NBA playoff series outcomes. The 2020 playoff provided a unique
opportunituy to study the effects of the bubble, and home-court
advantage by comparing the 2020 season with the seasons in the past
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games so scoring is the primary focus 
of this study. The specific measures of interest are team scoring totals and 
team shooting percentage on two pointers, three pointers, and free throws. 
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows that the 2020 playoffs favored away teams more than normal, particularly
with two point shooting and total scoring.

```{r setup, echo = FALSE, message =  FALSE, warning = FALSE}
library(BSDA)
library(ggplot2)
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, dpi = 300,
out.width = "90%", fig.align = "center", dev = "pdf", fig.pos = "H")
```

# Introduction  
Home-court advantage is often discussed in sports circles as a contributing 
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a 
bubble brought a great deal of concern for fans, teams, journalists, and others. 
Steve Aschburner discusses details about anticipated effects in his article 
“How Orlando's Neutral Site Will Impact Teams, Referees and Games” @Aschburner. 
Aschburner shares concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the "home” team logo on
the court and allowing the "home" team to play crowd noise and music, but most 
people doubted these small attempts would recreate a true playoff atmosphere. 
During the 2020 NBA playoffs, home teams only won about 48.2% of the games. This
is lower than normal, which Aschburner claims usually floats around 60%. This 
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.

@@

This study is quite different from earlier NBA home-court advantage studies. 
By using the neutral site games of 2020 we will get to compare home and away
performance to a control. Typically, studies just compare home vs away 
performance. These studies don’t separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a 
result of home teams overperforming or away teams underperforming because of 
home-court advantage. Some of these studies are reviewed in greater detail in
Section II Literature Review.

@@

# II. Literature Review  
There is tons of published literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki examines the impact of home-court advantage on winning in the
NBA. Kotecki proves the existence of home-court advantage using performance-based
statistics, specifically field goal percentage, free throw percentage, and 
points. All of which he proves significantly improved from home-court advantage, 
indicating home-court helps teams play better. This conclusion comes from 
comparing home performance vs. away performance in games. @Cao studied the effects
of pressure on performance in the NBA. They use free throws as their measure of 
interest, and research home and away factors in the study. They expected that home
fans could distract and put pressure on opposing players to make free throws,
which would negatively impact away team shooting. Likewise, the home team
would see a positive effect on free throw shooting. However, they find insignificant
evidence that home status has a substantial impact on missing from the free throw 
line. @Harris use two point shots, three-point shots and free throws as measures
of interest to study home-court advantage. They find that two point shots are
the strongest predictor of home-court advantage. They suggest that evidence 
supports that home teams should try to shoot more two point shots and force
their opponent to take more two point shot attempts. This strategy will maximize
the benefits of home-court advantage and give them the best chance to win.

@@

# III. Data and Methods  
The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few 
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history the bubble allows NBA home and away performance
to be compared vs a control/neutral field. The NBA bubble, as a neutral court,
removed all 4 possible game location factors impacting home-court advantage 
hypothesized by @Carron1992. The NBA bubble featured 8 seeding games then a
standard playoff format. The focus of this study was the play during the playoff
games since it followed the standard playoff format and can easily be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. It is important to make sure that any
observed differences are truly from a home-court advantage. So, in order to control
for the changing play style of the NBA, we will limit the study to 2020 vs 2017-2019.
The reason being the faster pace play and more common use of the three-point shot 
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very 
close in pace and playing style to 2020.

@@

# IV. Analysis and Results
```{r, echo=FALSE}
#Analysis
one <- prop.test(x=c(sum(game_log_2020_home$W.L), sum(game_log_2017_2019_home$W.L)), 
                 n=c(nrow(game_log_2020_home), nrow(game_log_2017_2019_home)), correct=FALSE)
rowone <- c(as.numeric(one$estimate[1]), as.numeric(one$estimate[2]), one$conf.int,
          sqrt(as.numeric(one$statistic)), one$p.value)

@@

# V. Discussion 
Generally it seemed that away teams fared better in the 2020 NBA playoff bubble
than previous years on the road. Starting from the dip in home winning percentage
to below .482 it’s clear something was different. Although the difference was not
significant after a Bonferroni correction it’s still important to consider and
understand that home teams seemed to struggle to win compared to normal conditions.
Compared to @Kotecki who finds home teams consistently have a significantly
better record than away teams boasting about a 60.5% win percentage in his sample,
the 48.2% home winning percentage of 2020 home teams is a shift from normal.
It’s clear that home teams did not benefit from the usual advantages provided by
being the home team.

@@

# References

>>>>

---
title: "The Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price and Jun Yan
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: bookdown::pdf_document2
header-includes:
   - \usepackage[margin=1in]{geometry}
toc: False
---
Abstract: 
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there would be no fans in attendance, 
games played on neutral courts and no traveling for teams, which in theory 
removes home-court advantage from the games. This setting has attracted a lot of
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played a huge role in 
NBA playoff series outcomes. The 2020 playoff provided a unique
opportunituy to study the effects of the bubble, and home-court
advantage by comparing the 2020 season with the seasons in the past
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games so scoring is the primary focus 
of this study. The specific measures of interest are team scoring totals and 
team shooting percentage on two pointers, three pointers, and free throws. 
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows that the 2020 playoffs favored away teams more than normal, particularly
with two point shooting and total scoring.

```{r setup, echo = FALSE, message =  FALSE, warning = FALSE}
library(BSDA)
library(ggplot2)
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, dpi = 300,
out.width = "90%", fig.align = "center", dev = "pdf")
```

# Introduction {#sec:intro}
Home-court advantage is often discussed in sports circles as a contributing 
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a 
bubble brought a great deal of concern for fans, teams, journalists, and others. 
Steve Aschburner discusses details about anticipated effects in his article 
“How Orlando's Neutral Site Will Impact Teams, Referees and Games” @Aschburner. 
Aschburner shares concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the "home” team logo on
the court and allowing the "home" team to play crowd noise and music, but most 
people doubted these small attempts would recreate a true playoff atmosphere. 
During the 2020 NBA playoffs, home teams only won about 48.2% of the games. This
is lower than normal, which Aschburner claims usually floats around 60%. This 
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.

@@

This study is quite different from earlier NBA home-court advantage studies. 
By using the neutral site games of 2020 we will get to compare home and away
performance to a control. Typically, studies just compare home vs away 
performance. These studies don’t separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a 
result of home teams overperforming or away teams underperforming because of 
home-court advantage. Some of these studies are reviewed in greater detail in
Section \@ref(sec:litrev) Literature Review.

@@

# Literature Review {#sec:litrev}
There is tons of published literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki examines the impact of home-court advantage on winning in the
NBA. Kotecki proves the existence of home-court advantage using performance-based
statistics, specifically field goal percentage, free throw percentage, and 
points. All of which he proves significantly improved from home-court advantage, 
indicating home-court helps teams play better. This conclusion comes from 
comparing home performance vs. away performance in games. @Cao studied the effects
of pressure on performance in the NBA. They use free throws as their measure of 
interest, and research home and away factors in the study. They expected that home
fans could distract and put pressure on opposing players to make free throws,
which would negatively impact away team shooting. Likewise, the home team
would see a positive effect on free throw shooting. However, they find insignificant
evidence that home status has a substantial impact on missing from the free throw 
line. @Harris use two point shots, three-point shots and free throws as measures
of interest to study home-court advantage. They find that two point shots are
the strongest predictor of home-court advantage. They suggest that evidence 
supports that home teams should try to shoot more two point shots and force
their opponent to take more two point shot attempts. This strategy will maximize
the benefits of home-court advantage and give them the best chance to win.

@@

# Data {#sec:data}
The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few 
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history the bubble allows NBA home and away performance
to be compared vs a control/neutral field. The NBA bubble, as a neutral court,
removed all 4 possible game location factors impacting home-court advantage 
hypothesized by @Carron1992. The NBA bubble featured 8 seeding games then a
standard playoff format. The focus of this study was the play during the playoff
games since it followed the standard playoff format and can easily be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. It is important to make sure that any
observed differences are truly from a home-court advantage. So, in order to control
for the changing play style of the NBA, we will limit the study to 2020 vs 2017-2019.
The reason being the faster pace play and more common use of the three-point shot 
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very 
close in pace and playing style to 2020.

@@

# Methods {#sec:methods}

# Analysis and Results {#sec:results}
```{r, echo=FALSE}
#Analysis
one <- prop.test(x=c(sum(game_log_2020_home$W.L), sum(game_log_2017_2019_home$W.L)), 
                 n=c(nrow(game_log_2020_home), nrow(game_log_2017_2019_home)), correct=FALSE)
rowone <- c(as.numeric(one$estimate[1]), as.numeric(one$estimate[2]), one$conf.int,
          sqrt(as.numeric(one$statistic)), one$p.value)

@@

# Discussion {#sec:disc}
Generally it seemed that away teams fared better in the 2020 NBA playoff bubble
than previous years on the road. Starting from the dip in home winning percentage
to below .482 it’s clear something was different. Although the difference was not
significant after a Bonferroni correction it’s still important to consider and
understand that home teams seemed to struggle to win compared to normal conditions.
Compared to @Kotecki who finds home teams consistently have a significantly
better record than away teams boasting about a 60.5% win percentage in his sample,
the 48.2% home winning percentage of 2020 home teams is a shift from normal.
It’s clear that home teams did not benefit from the usual advantages provided by
being the home team.

@@

# References {#sec:ref}
